\input{../preamble.tex}
\usepackage[margin=0.5in]{geometry}
\begin{document}
\title{Models of linear dependent type theory}
\author{Martin Lundfall}
\maketitle
\abstract
Linear dependent type theory is the synthesis of linear logic and dependent type theory, yielding a system of deduction which maintains a careful account of the assumptions used in a derivation. In this master thesis I explore some concrete models in which linear dependent type theory can be interpreted.
\newpage
\tableofcontents
\newpage
\section{Background}

\subsection{Dependent type theory}

\subsection{Linear type theory}
\subsection{Fibrations}

A general heuristic for modelling some kind of mathematical object $E$ dependent over some other object $B$ is to specify a \textit{projection} morphism $p : E \to B$ subject to certain constraints. For example, a family of sets $A_i$ indexed the set $I$ might equally well be understood as the set $E = \bigsqcup_{i \in I}A_i$ together with a morphism $p : E  \to I$ such that $p^{-1}(i) = A_i$. This is the principle behind fibrations. Dependeding on which kind of mathematical object we are dealing with, we want to impose conditions on the projection $p : E \to B$ to be able to \textit{lift} certain structure of the base $B$ into \textit{fibers} $E_b = p^{-1}(b)$ for $b \in B$.

\subsubsection{Grothendieck fibrations}
In the context of categories, the appropriate notion is that of a \textbf{Grothendieck fibration}, or simply \textbf{fibration}. In order for a functor $p : E \to B$ to be a fibration, one needs to be able to lift arrows in the base category $B$ to arrows in $E$. We do this by asking for the existence of certain \textit{cartesian arrows} in $E$.
\begin{defn}[Cartesian arrow]
Given a functor $p : E \to B$, an arrow $f : e' \to e''$ of $E$ is said to be \textbf{cartesian} with respect to $p$ if for every $h : e \to e''$ and $\alpha : p(e) \to p(e')$ such that $p(f)\alpha = p(h)$, there is a unique arrow $\hat \alpha : e \to e'$ such that $p(\hat \alpha) = \alpha$ and $f \hat \alpha = h$.
\end{defn}
The situation is illustrated in the following diagrams:
\[
\ti
p(e) \ar[d, "\forall \alpha"] \ar[dr, "p(h)"] & & & e \ar[d, "!\exists \hat \alpha"] \ar[dr, "\forall h"] \\
p(e') \ar[r, "p(f)"] & p(e'') & & e' \ar[r, "f"] & e''\\
\kz
\]
\begin{defn}[Grothendieck fibration]
A functor $p: E \to B$ is a \textbf{Grothendieck fibration} if, for every $e \in E$, $b \in B$ and arrow $f : b \to p(e)$ in $B$, there exists a cartesian arrow $f^*$ in $E$ such that $p(f^*) = f$.
\end{defn}
Sometimes we will refer to $f^*$ as a \textbf{lift} of $f$. Although it is not uniquely determined, its domain will determined up to unique isomorphism. 
A basic review of Grothendieck fibrations particularly relevant to the semantics of dependent type theory can be found in \cite{jacobs}. We repeat the notions relevant for our investigations here.
\\

For any object $b \in B$ in the base, we call the subcategory of $E$ which is mapped to $b$ and its identity morphism the \textbf{fiber} over $b$. This will be denoted $E_b$.
A fibration $p : E \to B$ induces for every $u : b \to b'$ in $B$ a functor $u^*: E_{b'} \to E_{b}$ by sending every object to the domain of the cartesian arrow associated with it. This functor will be unique up to unique natural isomorphism. In general, for two compatible morphisms $u$ and $v$ in the base, a lift of a composition is only naturally isomorphic to the composition of a lift. In other words, we have $u^* \circ v^* \cong (u \circ v)^*$ and not functoriality on the nose. Those fibration for which there are lifts of morphisms such that the equalities hold on the nose are called \textbf{Split}.

\begin{defn}[Cartesian functor]
Let $p : E \to B$ and $q : E' \to B$ be fibrations over the same base. A functor $F : E \to E'$ is \textbf{cartesian} if $qF = p$ and cartesian morphisms in $E$ with respect to $p$ are mapped to cartesian morphisms in $E'$ with respect to $q$.
\end{defn}
This determines a category $\textbf{Fib}(B)$, consisting of fibrations over $B$ and cartesian functors between them. More generally, one can construct a category $\textbf{Fib}$ whose morphisms from fibrations $p : E \to B$ and $q : E' \to B'$ are pairs of functors $(F, G)$ where $F : E \to E'$ and $G : B \to B'$ such that $G \circ p = q \circ F$ and $F$ preserves cartesian morphisms. In fact, the functor $\textbf{Fib} \to \textbf{Cat}$ sending a fibration to its base is itself a fibration whose fibers are $\textbf{Fib}(B)$ for any small category $B$.


With the notion of a fibered natural transformation, $\mathbf{Fib}$ forms a 2-category.
\begin{defn}[Fibered natural transformation]
For two pairs of parallel functors $F$, $H$ and $G$, $J$ between fibrations $(E, B)$ and $(E', B')$, as illustrated in the commutative square:
\[
\ti
E \ar[d, "p"] \ar[r, "F", bend left] \ar[r, "G", bend right] & E' \ar[d, "{p'}"]\\
B \ar[r, "H", bend left] \ar[r, "J", bend right] & B'
\kz
\]
a \textbf{Fibered natural transformation} $(\lambda, \lambda')$ between $(F, H)$ and $(G, J)$ is a pair of natural transformations $\lambda : G \to F$ and $\lambda' : A \to B$ such that $p'(\lambda) = p_{\lambda'}$.
\end{defn}
\begin{rem} Notice that this definition does not ask for $F$ and $G$ to be cartesian functors. But when that is the case, fibered natural transformations between the parallel morphisms $(F, A)$ and $(G, A)$ in $\mathbf{Fib}$ are the 2-morphisms of this 2-category.
\end{rem}
An important special case of this is when $B = B'$ and $H = J = 1_B$. Then a fibered natural transformation $\lambda : F \to G$ is simply a natural transformation such that all of its components are sent to identities via $p'$.


TODO: Put this in the right context
Let $\mathcal{C}$ be a category and $F : \mathcal{C} \to \mathbf{Cat}$ a (pseudo)functor. $\textbf{Cat}_{/*}$ denotes the slice category of categories over the terminal category, i.e. the category constisting of pointed categories $(\mathcal{D}, d)$, $(\mathcal{E}, e)$ as objects and functors $G : \mathcal{D} \to \mathcal{E}$ equipped with morphisms $\gamma : G(d) \to e$ as morphisms.
\begin{defn}
The Grothendieck construction $\int F$ of $F$ is the pullback of the following diagram:
\[
\ti
\int F \ar[r] \ar[d] & \textbf{Cat}_{/*} \ar[d, "U"] \\
\mathcal{C} \ar[r, "F"] & \textbf{Cat} \\
\kz
\]
\end{defn}
In other words, the category whose objects consists of pairs $(a, b)$ where $a \in \Ob(\mathcal{C})$ and $b \in F(a)$, and whose morphisms $f : (a, b) \to (\alpha, \beta)$ are pairs $(f, g)$ where $f: a \to \alpha$ and $g : F(f)(b) \to \beta$.

\subsubsection{Other types of fibrations}
In exploring models of linear and dependent types, fibrations of other structures will arise. In particular, we will need to talk about fibration of groupoids and lax monoidal fibrations.
\begin{defn}
A map $p : G \to H$ in $\textbf{Grpd}$ is a \textbf{fibration of groupoids} if for every $g \in G$
and $f : p(g) \to h$ in $H$, there exists an object $g'$ and map $\hat f : g \to g'$ in $G$ such that $p(g') = h$ and $p(\hat f) = f : p(g) \to p(g')$.
\end{defn}
When considering fibrations of monoidal categories, we distinguish between the case where both the fibration and the base are monoidal categories, and when we simply want each fiber to be a monoidal category and the induced functors between these to be monoidal functors. The former notion is that of a monoidal fibration:
\begin{defn}
A \textbf{monoidal fibration} is a functor $\Phi\colon E\to B$ such that
\begin{itemize}
\item $\Phi$ is a Grothendieck fibration
\item $E$ and $B$ are monoidal categories and $\Phi$ is a strict monoidal functor, and
\item the tensor product of $E$ preserves cartesian arrows.
\end{itemize}
\end{defn}
From this it can be shown that maps in the base induce monoidal functors in between the fibers. See \cite{shulmanmonoidal} for details.


A weaker notion that does not require the base category to be monoidal is that of a lax monoidal fibration \cite{zawadowski}. We simply want each fiber to be a monoidal category and that funcotrs between fibers are monoidal.
\begin{defn}
A \textbf{lax monoidal fibration} is a fibration $p : E \to B$ along with
\begin{enumerate}
\item Two functors $\otimes : E \times_{B} E \to E$ and $I : B \to E$ fitting into the following diagram:
\[
\ti
E \times_{B} E \ar[r, "\otimes"] \ar[rd] & E \ar[d, "p"] & B \ar[l, "I"] \ar[ld, "1_B"] \\
& B &
\kz
\].
\item Three fibered natural isomorphisms $\alpha, \lambda$ and $\rho$ associated with the diagrams:
\[
\ti
E \times_B E \times_B E \ar[r, "1_E \times_B \otimes"] \ar[d, "\otimes \times_B 1_E"] & E \times_B E \ar[d, "\otimes", swap] \\
E \times_B E \ar[r, "\otimes"] \ar[ru, "\alpha", Rightarrow, shorten <=30pt,shorten >=30pt]  & E
\kz
\]
and
\[
\ti
B \times_B E \ar[r, "I \times_B 1_{E}"] \ar[rdd, "\pi_2",swap] & E \times_B E  \ar[dd, "\otimes",swap] & E \times_B B \ar[ldd, "\pi_1"] \ar[l, "1_E \times I"] \\
\ar[r, "\lambda", xshift=20 pt, Rightarrow, shorten <=30pt, shorten >=30pt] & {}& \ar[l, "\rho", Rightarrow, xshift=-20pt, shorten <=30pt, shorten >=30pt,swap] \\
  & E &  
\kz
\]
\item such that $\alpha$, $\lambda$ and $\rho$ satisfies the pentagon and triangle identities in each fiber.
\item for every $b \in B$, $\rho_{I_b} = \lambda^{-1}_{I_b} : I_b \otimes I_b \to I_b$
\end{enumerate}
\end{defn}
These conditions are sufficient for each fiber to be a monoidal category and for the induced functors between fibers to be monoidal. More detais are given in \cite{zawadowski}.
\subsection{Spectra}
A sequential spectrum is an $\mathbb{N}-graded$ (compactly generated, or (sufficiently well-behaved)) pointed topological space $X_{\bullet} = X_n$ with continuous maps $X_n \land S \to X_{n+1}$. It is not uncommon that the maps (often referred to as structure maps) are in fact isomorphisms.
A basic example is the suspension spectra, which can be constructed out of any (well-behaved) topological space $X$ as $X_n = S^n \land X$. The map $S^n \land X \land S \to S^{n+1} \land X$ is an isomorphism.
Every spectra corresponds to a cohomology theory, as outlined below.
In Homotopy type theory, Mike Shulman writes in (https://homotopytypetheory.org/2013/07/24/cohomology/), spectrums are described as follows:
More precisely, a spectrum is a family of types $Y : \mathbb{N} \to \mathcal{U}$ such that for all n we have (a specified equivalence) $Y_n = \Omega Y_{n+1}$.
For such a Y, we define the (abelian) cohomology of X with coefficients in Y to be
\[
H^n(X;Y) := \Vert X \to \Omega^{k-n} Y_k \Vert_0 \quad (n \in \mathbb{Z})
\]
The stable homotopy group of a spectra is a $\mathbb{Z}$-graded abelian group, whose $q$ component arises as a colimit to the diagram:
\[
\pi_{k+q}(X_k) = [S^{k+q}, X_k]_* \xrightarrow{S \land (-)} [S^{k+q+1}, S \land X_{k}] \xrightarrow{1_{S^{k+q+1}}, \sigma} [S^{k+q+1}, X_{k+1}] = \pi_{k+q+1}(X_{k+1})
\]
Definition 1.16. A morphism $f: X \to Y$ of sequential spectra, is called a stable weak homotopy equivalence, if its image under the stable homotopy group-functor of def. 1.13 is an isomorphism
\[
\pi_{\bullet}(f): \pi_{\bullet}(X) \simeq \pi_{\bullet}(Y)
\]
An $\Omega$-spectrum is a spectrum for which the maps $\hat \sigma : X_n \to \text{Maps}(S^1,X_{n+1})$ arising out out of the adjunction $\Hom(\Sigma X, Y) \cong \Hom(X, [S^1, Y])$ from the structure maps $\sigma$ are weak homotopy equivalences. Note that this is the condition Mike requires above.
The category of spectra has all limits and colimits. They are inherited from Top. It also has a zero object, the spectra which is a constant point. Coproducts of spectra $X$ and $Y$ comes from the componentwise wedge sum of topological spaces.
A map $f: A \to B$ is called a \textbf{retract} (of $g : C \to D$) if there exist i, j, r, and s, such that the following diagram commutes:
\[
\ti
A \ar[d, "f"] \ar[r, "i"] & C \ar[d, "g"] \ar[r,"j"] & A \ar[d, "f"]\\
B \ar[r, "r"] & D \ar[r, "s"] & B
\kz
\]
such that $j \circ i = 1_A$ and $s \circ r = 1_B$.
\begin{defn}
A complete and cocomplete category is a \textbf{model category}, $\mathcal{M}$ if it has three classes of morphisms, F, G and W called fibrations, cofibrations and weak equivalences satisfying the following conditions:
\begin{itemize}
\item \textbf{Retracts}: If $f$ is a retract of $g$, then $f$ belongs to the same class as $g$.
\item \textbf{(2 out of 3)}: If two of the maps $f$, $g$ and $gf$ are weak equivalences, then so is the third
\item \textbf{Lifting} (weak equivalences + cofibrations) have the left lifting property with respect to fibrations, and cofibrations have the left lifting property with respect to (weak equivalances + fibrations). In other words, for $f$ a fibration and $g$ cofibration, if anyone of them is also weak equivalence, then there exists a lifting map as in the following diagram:
\[
\ti
A \ar[d, "f"] \ar[r] & B \ar[d, "g"]\\
C \ar[ur, "h", dotted] \ar[r] & D
\kz
\]
\item \textbf{Factorization} every morphism f can be written as $f = p \circ i$ for a fibration p and an acyclic cofibration i. Every morphissm f can be written as $f = p \circ i$ for an acyclic fibration p and a cofibration i.
\end{itemize}
\end{defn}
An object is said to be fibrant if $X \to 1$ is a fibration. An object is cofibrant if $0 \to X$ is a cofibration.\\
A \textbf{path object} $PX$ in a model category is the factorization object in the diagonal map $\Delta_X : X \to X \times X$.
\[
X \xrightarrow{i \in W} PX \xrightarrow{p \in F} X \times X
\]
A \textbf{cylinder object} $CX$ is the factorization of the codiagonal map $\nabla_X : X \sqcup X \to X$:
\[
X \sqcup X \xto{f \in G} CX \xto{g \in W} X
\]
The factorization axiom ensures that these always exists (in fact, factorization is a stronger condition)\\
The fiber of a map of pointed spaces $f : X \to Y$ is the pullback:
\[
\ti
\text{fib}(f) \ar[r] & X \ar[d, "f"]\\
* \ar[r, "point"] & Y
\kz
\]
where point is the inclusion of the point into the selcted point of $Y$. Cofiber arises dually as a pushout.
For $f, g : X \to Y$, a \textbf{left homotopy} $\eta : f \implies_L g$ is a map $\eta : CX \to Y$ such that the following diagram commutes:
\[
\ti
X \ar[dr, "f"] \ar[r] & CX \ar[d, "\eta"] & X \ar[l] \ar[dl, "g",swap]\\
& Y &
\kz
\]
For $f, g : X \to Y$, a \textbf{right homotopy} $\eta : f \implies_R g$ is a map $\eta :X \to PY$ such that the following diagram commutes:
\[
\ti
& X \ar[dl, "f"] \ar[dr, "g"] \ar[d, "{\eta}"]\\
Y &   PY \ar[l,] \ar[r, ] & Y
\kz
\]
If $X$ is a cofibrant object, then a left homotopy induces a right homotopy. If $X$ is fibrant, then a right homotopy induces a left homotopy.
The \textbf{Homotopy category}, $Ho(\mathcal{C})$ of a model category $\mathcal{C}$ is the category whose objects are the objects of $\mathcal{C}$ which are both fibrant and cofibrant and whose morphisms are the equivalence classes of the morphisms of $\mathcal{C}$ under the homotopy relation.\\
From any map $f: X \to Y$ one can form long fiber sequences:
\[
\dots \text{fib(fib(f))} \cong \Omega Y \to \text{fib}(f) \to X \xto{f} Y
\]
or long cofiber sequences analoguously.
Spectra form a model category. A homomorphism of spectra $f_{\bullet} : X_{\bullet} \to Y_{\bullet}$ is a
\begin{itemize}
\item \textbf{Strict weak equivalence} if the components $f_n : X_n \to Y_n$ are weak homotopy equivalences.
\item \textbf{strict fibration} if each component is a fibration in the classical model structure on topological spaces. (Serre fibration)
\item \textbf{strict cofibration} if $f_0$ is a cofib in Top, and if all maps
\[
f_{n+1}, \sigma : X_n \sqcup_{S \land X_{n}} S \land Y_{n} \to Y_{n+1}
\]
\end{itemize}
are cofibrations.\\
These classes of morphisms give the category of spectra the structure of a model categoyr.\\
A \textbf{relative cell complex inclusion} is a map $f : X \to X'$ fitting in to the following pushout diagram:
\[
\ti
S^{n-1} \ar[r,"\phi"] \ar[d, "i"] & X \ar[d,"f"]\\
D^n \ar[r] & X'
\kz
\]
for some continuous function $\phi$.\\
A spectrum is called a CW-spectrum if all of its spaces are CW-complexes and the structure maps are relative cell complex inclusions.\\
The category of spectra can be equipped with fibrations, cofibrations and weak equivalences in such a way that the fibrant-cofibrant objects are all $\Omega$-spectra. The homotopy category of this model category is called the \textbf{stable homotopy category}.\\
The stable homotopy category has finite colimits induced by the wedge sum, and zero object $0=\sum(\star)$. The hom-sets $[X, Y]$ of the stable homotopy category have a group structure, since $[X, Y] \cong [X, \Omega \Sigma Y]$ and since maps into a loop object carry an induced group action given by path composition. Furthermore this group is abelian (Eckmann-Hilton). Thus the stable homotopy category is a category enriched in abelian groups.
An additive category is a category $\mathcal{C}$:
\begin{itemize}
\item Enriched in \textbf{Ab}
\item with finite coproducts
\end{itemize}
from this follows that the coproducts coincide with products.\\
The stable homotopy category is therefore an additive category.
It is furthermore a triangulated category, which makes the long homotopy fiber sequences and cofiber sequences coincide. Specifically, for a cofiber sequence $X \xto{f} Y \xto{g} Z$ in the stable homotopy category and for any other spectra $A$, there is a long \textbf{fiber-cofiber} sequence of abelian groups:
\[
\dots [A, \Omega Y] \to [A, \Omega Z] \to [A, X] \to [A, Y] \to [A, Z] \to [A, \Sigma X] \to [A, \Sigma Y] \dots
\]
\newpage

\section{Syntax}
The particular linear dependent type theory under consideration is a mixture of the presentations given by Krishnaswami in \cite{krishnaswami} and Vákár in \cite{vakar14}. As is common in linear logic, we split contexts into two parts separated by a semicolon. The first part contains \textit{intuitionistic} assumptions who can be used freely (or not used at all) in the conclusion and are not 'consumed' when used. The second part of the context is the \textit{linear} part, containing ephemeral assumptions or resources, who have to be used exactly once in the conclusion. Types can either be intuitionistic, in which case we simply write $A \text{ type}$, or linear, written $A \text{ linear}$. 
\subsection{Structural rules}
We will be dealing with the following judgements:\\
\fbox{
\begin{minipage}{.3\linewidth}
\textbf{Judgement:}\\
$\vdash \Gamma \text{ ctxt}$\\
$\vdash \Gamma; \Xi \text{ ctxt}$\\
$\Gamma \vdash A \text{ type}$\\
$\Gamma \vdash A \text{ linear}$\\
$\Gamma \vdash M : A$\\
$\Gamma; \Xi \vdash x : A$\\
$\Gamma \vdash A \equiv A' \text{ type}$\\
$\Gamma \vdash A \equiv A' \text{ linear}$\\
$\Gamma \vdash M \equiv N : A$\\
$\Gamma; \Xi \vdash x \equiv y : A$
\end{minipage}
\begin{minipage}{.7\linewidth}
\textbf{Meaning:}\\
$\Gamma$ is a well-formed intuistionistic context.\\
$\Gamma; \Xi$ is a well-formed mixed context\\
$A$ is an intuistionistic type in $\Gamma$\\
$A$ is a linear type in $\Gamma$\\
$M$ is a term of the intuitionistic type $A$ in $\Gamma$\\
$x$ is a (linear) term of the linear term of $A$ in $\Gamma; \Xi$\\
$A$ and $A'$ are equal intuitionistic types in $\Gamma$\\
$A$ and $A'$ are equal linear types in $\Gamma; \Xi$\\
$M$ and $N$ are equal intuitionistic terms of $A$ in $\Gamma$\\
$x$ and $y$ are equal linear terms of $A$ in $\Gamma; \Xi$
\end{minipage}}
\\

Our basic structural rules are the following:
\\
\fbox{
\begin{minipage}{.5\linewidth}
\[
\inference {}{\vdash \cdot \text{ ctxt}}[CI-Base]
\]
\[
\inference {}{\vdash \cdot ; \cdot \text{ ctxt}}[CM-Base]
\]
\[
\inference {\Gamma \vdash A \text{ type}}{\vdash \Gamma, x : A \text{ ctxt}}[C-int-ext]
\]
\[
\inference{\Gamma \vdash A \text{ type} \quad \vdash \Gamma, \Delta \text{ ctxt}}{
\vdash \Gamma, x : A, \Delta \text{ ctxt}}[C-weak-1]
\]
\[
\inference{\Gamma \vdash A \text{ type} \quad \vdash \Gamma, \Delta; \Xi \text{ ctxt}}{
\vdash \Gamma, x : A, \Delta; \Xi \text{ ctxt}}[C-weak-2]
\]
\[
\inference {\vdash \Gamma; \Xi \text{ ctxt} \quad \Gamma \vdash A \text{ linear}}{\vdash \Gamma; \Xi, x : A \text{ ctxt}}[C-lin-ext]
\]
\[
\inference{\Gamma; \Xi, x : A, y : B, \Xi' \vdash t : A'}
{\Gamma; \Xi, y : B, x : A, \Xi' \vdash t : A'}[Lin-exch]
\]
\end{minipage}
\begin{minipage}{.5\linewidth}
\[
\inference{\Gamma \vdash A \text{ type} \quad \Gamma, \Gamma'; \Xi \vdash t : A'} {\Gamma, x : A, \Gamma'; \Xi \vdash t : A'}[Weak-L]
\]
\[
\inference{\Gamma \vdash A \text{ type} \quad \Gamma, \Gamma' \vdash \mathcal{J}} {\Gamma, x : A, \Gamma' \vdash \mathcal{J}}[Weak-I]
\]
\[
\inference{\Gamma \vdash M : A \quad \Gamma, x : A, \Gamma' \vdash \mathcal{J}} {\Gamma, M : A, \Gamma'[M/x] \vdash \mathcal{J}[M/x]}[Int-subst-1]
\]
\[
\inference{\Gamma \vdash M : A \quad \Gamma, x : A, \Gamma'; \Xi \vdash t : A'} {\Gamma, M : A, \Gamma'[M/x]; \Xi[M/x] \vdash t : A'[M/x]}[Int-subst-2]
\]
\[
\inference{\Gamma; \Xi, x : A \vdash t : B \quad \Gamma; \Xi' \vdash M : A}
{\Gamma; \Xi, \Xi' \vdash t[M/x]: B}[Lin-subst]
\]
\[
\inference{\Gamma, x : A, \Gamma' \text{ ctxt}}{\Gamma, x : A, \Gamma' \vdash x : A}[Int-var]
\]
\[
\inference{\vdash \Gamma; x : A \text{ ctxt}}{\Gamma; x : A \vdash x : A}[Lin-var]
\]
\end{minipage}
}
along with rules specifying that judgemental equality is closed over these. $\mathcal{J}$ denotes a judgement which is either $A \text{ type}$, $A \text{ linear}$ or $M : A$.
\subsection{Purely Intuitionistic typing}
Standard $\Pi$, $\Sigma$ and $Id$-types.
\subsection{Purely Linear typing}
Standard $\otimes$, $I$, $\multimap$, \&.
\subsection{Linear-intuitionistic interplay}
We want to have a way of types and terms back and forth between the intuitionistic and linear side. On the syntactic side, we introduce two the operators $M$ and $L$, which transfers a linear type/term to its intuitionistic counterpart and vice versa. Semantically, this will establish a fiberwise adjunction between the categories of linear types intuitionistic types:
\[
\ti
\mathcal{L}_{\Gamma} \ar[r, "M"{name=A, below}, bend right] & \mathcal{T}_{\Gamma} \ar[l, "L"{name=B, above}, bend right] \ar[from=A, to=B, symbol=\vdash]
\kz
\]
where the exponential modality from traditional linear logic is understood as the comonad $! = LM$. Below are the syntactic rules for the operators $M$ and $L$.\\
\fbox{
\begin{minipage}{.5\linewidth}
\[
\inference{\Gamma \vdash A \type}{\Gamma \vdash A_L \linear}[L-F]
\]\\
\[
\inference{\Gamma \vdash a : A}{\Gamma ; \cdot \vdash a_L : A_L}[L-I]
\]\\
\[
\inference{\Gamma \vdash B \linear \quad \Gamma; \Xi \vdash y : A_L \quad \Gamma, x : A; \Xi' \vdash t : B}{\Gamma; \Xi, \Xi' \vdash \text{let  $x$ be $y$ in }t : B}[L-E]
\]\\
\[
\inference{\Gamma; \Xi, \Xi' \vdash \text{let  $x$ be $F(s)$ in }t : B}{\Gamma; \Xi, \Xi' \vdash \text{let  $x$ be $F(s)$ in }t \equiv t[F(s)/x]  : B}[L-C]
\]
\end{minipage}
\begin{minipage}{.5\linewidth}
\[
\inference{\Gamma \vdash B \linear}{\Gamma \vdash B_M \type}[M-F]
\]\\
\[
\inference{\Gamma ; \cdot \vdash b : B}{\Gamma \vdash b_M : B_M}[M-I]
\]\\
\[
\inference{
\Gamma \vdash t : B_M\\
}{\Gamma; \cdot \vdash \sigma(t) : B \\}[M-E]
\]\\
\[
\inference{
\Gamma \vdash b_M : B_M\\
}{\Gamma; \cdot \vdash \sigma(b_M) \equiv b : B \\}[M-C]
\]
\end{minipage}
}\\


Notice that the elimination and computation rules for $M$ suggests that $\sigma$ is a section to $M$. In particular, this implies that $M$ is faithful functor which is injective on objects.

\subsubsection{Linear dependent types}
The linear variants of the intuitionistic $\Pi$- and $\Sigma$-types will be denoted $\sqcap$ and $\exists$, respectively.\\
\fbox{
\begin{minipage}{.5\linewidth}
\[
\inference{\Gamma \vdash A \type \quad \Gamma, x : A \vdash B \linear}{\Gamma \vdash \sqcap_{x : A}B \linear}[$\sqcap$-F]
\]\\
\[
\inference{\Gamma, x : A; \Xi \vdash b : B}
{\Gamma; \Xi \vdash \lambda x. b : \sqcap_{x: A}B}[$\sqcap$-I]
\]\\
\[
\inference{\Gamma; \Xi \vdash t : \sqcap_{x : A}B \quad \Gamma \vdash a : A}
{\Gamma; \Xi \vdash t(a) : B[a/x]}[$\sqcap$-E]
\]\\
\[
\inference{\Gamma; \Xi \vdash \lambda x. b (a) : \sqcap_{x : A}B}
{\Gamma; \Xi \vdash \lambda x. b (a) = b[a/x] : B[a/x]}[$\sqcap$-C]
\]
\end{minipage}
\begin{minipage}{.5\linewidth}
\[
\inference{\Gamma \vdash A \type \quad \Gamma, x : A \vdash B \linear}{\Gamma \vdash \exists_{x : A}B \linear}[$\exists$-F]
\]\\
\[
\inference{\Gamma \vdash s : A \quad \Gamma; \Xi \vdash b : B[s/x]}{\Gamma \vdash (s, b) : \exists_{x : A}B}[$\exists$-I]
\]\\
\[
\inference{\Gamma; \Xi \vdash t : \exists_{x : A}B \quad \Gamma, x : A; \Xi', y : B \vdash c : C}{\Gamma; \Xi, \Xi' \vdash \text{let $x, y$ be $t$ in c : C}}[$\exists$-E]
\]\\
\[
\inference{\Gamma; \Xi \vdash \text{let $x, y$ be $(s, t)$ in c : C}}
{\Gamma; \Xi \vdash \text{let $x, y$ be $(s, t)$ in c} \equiv c[s/x][t/y]}[$\exists$-C]
\]

\end{minipage}
}\\

\section{Semantics}
\subsection{Contextual categories of linear dependent types}
To interpret linear dependent type theory we begin by building a \textbf{contextual category}, an algebraic structure which abstracts the key features of the syntax. The idea is to construct the most general framework in which one can interpret the structural rules to deal with these once and for all. Once this has been done we can consider the extra conditions have to be imposed in order for the model to support various type constructors, and provide concrete models that satisfy these conditions.\\

In our case, this is only almost true. We will build a contextual category in which not only the structural rules are assumed, but also the rules for the tensor product of linear types. This simplification allows us to formulate the semantic structure in terms of symmetric monoidal categories instead of multicategories.

Our particular choice of contextual category will consist of a comprehension category \cite{jacobs}, in which we can interpret the intuitionistic fragment of our theory and a locally symmetric monoidal fibration where the linear fragment will reside.
\begin{defn}
A \textbf{comprehension category} consists of a commutative diagram of functors
\[
\ti
E \ar[rd, "p"] \ar[r, "\pi"]  & B^{\to} \ar[d, "\text{cod}"]\\
& B
\kz
\]

where $B^\to$ is the arrow category of $B$ and $cod : B^\to \to B$ denotes the codomainfunctor such that:
\begin{enumerate}%
\item $B$ has a terminal object
\item $p : E\to B$ is a Grothendieck fibration,
\item $\pi : E\to B^\to$ takes cartesian morphisms in $E$ to cartesian morphisms in $B^\to$
\end{enumerate}
Notice that by the second condition, cartesian morphisms in $E$ are mapped to pullback squares in $B$:\\
A cartesian morphism $(p, q) : f \to g$ in $B^\to$ is a commutative square in $B$
\[
\ti
B \ar[r, "g"] \ar[d, "p"] & B' \ar[d, "q"]\\
A \ar[r,"f"] & A'
\kz
\]
such that for any $E'$, $E$, $q' : E' \to B'$, $e : E \to E'$ and $p' : E \to A$ as in the following diagram:
\[
\ti
E \ar[d, "id_E"] \ar[r, "e"] & E' \ar[d, "q'"] \\
E \ar[d, "p'"] \ar[r, "q'e"] & B' \ar[d,"q"] \\
A \ar[r, "f"] & A'
\kz
\]
there is a unique arrow $u : E \to B$ such that $p' = p \circ u$ and $q'e = g \circ u$. This is precisely the universal property of the pullback.
If $\mathcal{B}$ has all pullbacks, then $\text{cod} : B^\to \to B$ is a fibration and in that case $\pi$ is a cartesian functor.
\begin{lem}\label{pb-lem}
The pullback of a section is the section of a pullback. In other words, for morphisms $f : A \to C$, $g : B \to C$ and $h : C \to B$ such that $hg = 1_B$, then $h^*$ is a section of $g^*$.
\begin{proof}
We have the following pullback:
\[
\ti
A \times_C B \ar[d, "q"] \ar[r, "g^*"] & A \ar[d, "f"] \\
B \ar[r, "g"] & C
\kz
\]
Pulling back $h$ along $q$ we have the following double pullback:
\[
\ti
E \ar[d, "q'"] \ar[r, "h^*"] & A \times_C B \ar[d, "q"] \ar[r, "g^*"] & A \ar[d, "f"] \\
C \ar[r, "h"] & B \ar[r, "g"] & C
\kz
\]
The two pullbacks lemma says that if the two inner squares are pullbacks, then so is the outer rectangle, so $E$ is also the pullback of the span $f : A \to C$ and $gh = 1_C : C \to C$. But this is just $A$ with projections $1_A$ and $f$. So we get that $q' = f$ and $g^* h^* = 1_A$.
\end{proof}
\end{lem}
\end{defn}
A semantics for an intuitionistic linear dependent type theory consists of a comprehension $p : \mathcal{T} \to \mathcal{C}$ and a lax monoidal fibration $q : \mathcal{L} \to \mathcal{C}$, illustrated in the following picture:
\[
\ti
\mathcal{L} \ar[rd, "q"] & \mathcal{T} \ar[d, "p"] \ar[r, "\pi"] & \mathcal{C^\rightarrow} \ar[ld, "\text{cod}"]\\
& \mathcal{C}
\kz
\]
In this structure, we will interpret intuitionistic contexts $\Gamma$ as objects of $\mathcal{C}$, linear types in $\Gamma$ as objects of the fiber $\mathcal{L}_{[[\Gamma]]}$ and intuitionistic types as objects in $\mathcal{T}_{[[\Gamma]]}$. Terms of linear types will be morphisms in the fibers of $\mathcal{L}$ and terms of intuitionistic types will be represented by a section of the projections associated with context extensions.\\
The correspondence between this structure and the structural rules are given by the following:
\begin{itemize}
\item CI-Base is the terminal object $\mathbf{1}$ in $\mathcal{C}$.
\item CM-Base is the unit object of $\mathcal{L}_{\mathbf{1}}$.
\item C-int-ext says that for any $A$ in $\mathcal{T}_{\Gamma}$, there is another object $\Gamma.A$ in $\mathcal{C}$. This object is the domain of the morphism that $A$ is mapped to via $\pi$:
\[\Gamma.A \xrightarrow {\pi_A} \Gamma\]
\item C-weak-1 corresponds to lifting the morphism $\pi_A : \Gamma.A \to \Gamma$ to a cartesian morphism $\pi_A^* : \mathcal{T}_\Gamma \to \mathcal{T}_{\Gamma.A}$. The context we are looking for is the one associated to $\pi_A^*(\Delta)$.
\item For C-weak-2, notice that since $\pi$ is a cartesian functor, the context above fits into the following pullback:
\[
\ti
\Gamma.A.\Delta\{\pi_A\} \ar[r, "q"] \ar[d, "\pi_{\Delta\{\pi_A\}}"]  & \Gamma.\Delta \ar[d, "\pi_\Delta"]\\
\Gamma.A \ar[r, "\pi_A"] & \Gamma
\kz
\]
so by lifting $q$ to the monoidal functor $q^* : \mathcal{L}_{\Gamma.\Delta} \to \mathcal{L}_{\Gamma.A.\pi^*(\Delta)}$. The context we are looking for is the image of $\Xi$ under this functor.
\item C-lin-ext says that for any two objects $\Xi$ and $A$ in $\mathcal{L}_{\Gamma}$, their tensor product $\Xi \otimes A$ is also an object of $\mathcal{L}_{\Gamma}$.
\item Int-exch says that if we have both $A$ and $B$ in $\mathcal{T}_{\Gamma}$, then $\Gamma.A.B\{\pi_A\}$ and $\Gamma.B.A\{\pi_B\}$ both are pullbacks of this cospan:
\[
\ti
& \Gamma.B \ar[d, "\pi_B"]\\
\Gamma.A \ar[r, "\pi_A"] & \Gamma
\kz
\]
since the cartesian lifts of these arrows are sent to pullback squares in $\mathcal{C}^\to$. Hence they are isomorphic, and so the object $\Gamma'$ in $\mathcal{T}_{\Gamma.A.B\{\pi_A\}}$ may equally well be thought of as an object in $\mathcal{T}_{\Gamma.B.A\{\pi_B\}}$. (Should probably make this more explicit).
\item Lin-exch is motivated since $\Xi \otimes A \otimes B \otimes \Xi' \cong \Xi \otimes B \otimes A \otimes \Xi'$ in $\mathcal{L}_{\Gamma}$.
\item Weak-L says that for any $A \in \mathcal{T}_{\Gamma}$ and morphism $t : \Xi \to A$ in $\mathcal{L}_{\Gamma, \Gamma'}$, we can transfer $t$ along the functor $q_{A, \Gamma}^* : \mathcal{L}_{\Gamma, \Gamma'} \to \mathcal{L}_{\Gamma.A.\Gamma'\{\pi_A\}}$ induced by the map $q_{A, \Gamma'}$ arising from the following pullback diagram:
\[
\ti
\Gamma.A.\Gamma'\{\pi_A\} \ar[r] \ar[d, "q_{A, \Gamma'}"]  & \Gamma.\Gamma' \ar[d, "\pi_\Gamma'"]\\
\Gamma.A \ar[r, "\pi_A"] & \Gamma
\kz
\]
to yield a morphism $q_{A, \Gamma'}^*(t) : q_{A, \Gamma'}^*(\Xi) \to q_{A, \Gamma'}^*(A')$.
\item Weak-I says that the morphism $q_{A, \Gamma'}$ above also induces a functor $q_{A, \Gamma'}^* : \mathcal{T}_{\Gamma, \Gamma'} \to \mathcal{T}_{\Gamma.A.\Gamma'.\{\pi_A\}}$.
\item For Int-subst-1, the judgement $\mathcal{J}$ can take three forms:
\[
\mathcal{J} = B[M/x] \text{ type}
\]
\[
\mathcal{J} = b[M/x] : B[M/x]
\]
\[
\mathcal{J} = B[M/x] \text{ linear}
\]
The categorical structure motivating these judgement all rely on the fact substitution in the type theory corresponds to pullbacks in $\mathcal{C}$. We begin with showing that there is a type $B[M/x]$ in the context $\Gamma. \Gamma'[M/x]$.\\
Suppose we are given the following objects of $\mathcal{T}$:
\[
\begin{split}
&A \in \mathcal{T}_{\Gamma}\\
&\Gamma' \in \mathcal{T}_{\Gamma.A}\\
&B \in \mathcal{T}_{\Gamma.A.\Gamma'}\\
\end{split}
\]
and sections:
\[
\begin{split}
&M : \Gamma \to \Gamma.A\\
\end{split}
\]
of the projection $\pi_A$. We first lift $M$ to a cartesian arrow ${M^*} : \Gamma.\Gamma'\{M^*\} \to \Gamma'$, which is sent to the following pullback square by $\pi$:
\[
\ti
\Gamma.\Gamma'\{M^*\} \ar[d, "\pi_{\Gamma'\{M^*\}}"] \ar[r, "q_{(M, \Gamma')}"] & \Gamma.A.\Gamma' \ar[d, "\pi_{\Gamma'}"]\\
\Gamma \ar[r,"M"] & \Gamma.A
\kz
\]
where ${q_{(M, \Gamma')}}$ is some map whose existance is given by the structure of the comprehension. We then lift $q_{(M, \Gamma')}$ to obtain a cartesian arrow $q_{(M, \Gamma')}^* : \Gamma.\Gamma'\{M^*\}.B\{q_{(M, \Gamma')}^*\} \to \Gamma.A.\Gamma'.B$ which fits in the following pullback diagram:
\[
\ti
\Gamma.\Gamma'\{M^*\}.B\{q_{(M, \Gamma')}^*\} \ar[d, "\pi_{B\{q_{(M, \Gamma')}^*\}}"] \ar[r,"q_{(q_{(M, \Gamma')},B)}"] & \Gamma.A.\Gamma'.B \ar[d, "\pi_B"] \\
\Gamma. \Gamma'\{M^*\} \ar[r, "q_{(M, \Gamma')}"]  & \Gamma.A.\Gamma'
\kz
\]
The element $\Gamma.\Gamma'\{M^*\}.B\{q^*_{M,\Gamma'}\}$ of $\mathcal{T}_{\Gamma.\Gamma'\{M^*\}}$ along with its associated projection will be our interpretation of $\Gamma.\Gamma'[M/x] \vdash B[M/x]$.\\
Now suppose there is a section $b : \Gamma.A.\Gamma' \to \Gamma.A.\Gamma'.B$ of the projection $\pi_B$. To display an element of $B[M/x]$ is to give a section of $\pi_{B\{q^*_{(M,\Gamma')}\}}$. By lemma \ref{pb-lem}, we get such a section by pulling back $b$ along $q_{(q_{(M, \Gamma')},B)}$.
\\
Finally, if $B$ is an object of $\mathcal{L}_{\Gamma.A.\Gamma'}$, then the image of $B$ under the functor $q_{(M, \Gamma')}^* : \mathcal{L}_{\Gamma.A.\Gamma'} \to \mathcal{L}_{\Gamma.\Gamma'\{M^*\}}$ will be our interpretation of $B[M/x]$ as a linear type in the context $\Gamma,\Gamma[M/x]$.
\item The interpretation of Int-subst-2 is the image of $t$ under $q_{(M, \Gamma')}^*$.
\item Lin-subst. Given morphisms $t : \Xi \otimes A \to B$ and $M : \Xi' \to A$ we get a morphism $t \circ (id_\Xi \otimes M) : \Xi \otimes \Xi' \to B$
\item Int-var 
\item Lin-var is given by the identity morphism on $A$ in $\mathcal{L}_{\Gamma}$.
\end{itemize}
\newpage
\section{Models}
\subsection{Set indexed families}
Our first model will be an extension of the standard set-theoretic model of dependent type theory \cite{hofmann1997syntax}. The extension of this model to the linear realm is fairly straightforward, but still provides a setting for some fairly interesting examples, such as $\mathcal{V}$-enriched categories. It is also sufficient for showing that linear dependent type theory is a proper generalisation of both dependent type theory and linear type theory, as outlined in \cite{vakar14}. 


Both the intuitionistic and linear fibrations will be categories of set-indexed families.
\begin{defn}[$Fam(\mathcal{C})$]
For an arbitrary category $\mathcal{C}$, let $Fam(\mathcal{C})$ denote the category whose objects consists of pairs $(S, f)$ where $S$ is a set and $f$ is a function $f : S \to Ob(\mathcal{C})$. Morphisms of $Fam(\mathbf{Set})$ are pairs $(u, \alpha) : (S, f) \to (S', g)$ where $u : S \to J$ and $\alpha : S \to \text{Mor}(\mathcal{C})$ such that $\alpha(s) : f(s) \to g(u(s))$ for all $s \in S$. Composition is given, for two compatible morphisms $(u, \alpha)$ and $(v, \beta)$, by:
\[
(v, \beta) \circ (u, \alpha) = (v \circ u, g(u(s)) \circ f(s))
\]
for all $s \in S$.
\end{defn}
By projecting a family to its indexing set, we get a fibration $p : Fam(\mathcal{C}) \to \mathbf{Set}$, and if $\mathcal{C}$ has a terminal object, $\top$, we can form a comprehension category by defining $\pi : Fam(\mathcal{C}) \to \mathbf{Set}^\to$ as follows.

On objects, let
\[
\pi(S, f) = fst: \{(s, t) \; | \; s \in S, t : \top \to f(s)\} \to S
\]
For a morphism $(u, \alpha) : (S, f) \to (S', g)$, let
\[
q_{(u,\alpha)} : \{(s, t) \; | \; s \in S, t : \top \to f(s) \} \to \{(s', t') \; | \; s' \in S', t' : \top \to g(s') \}
\]
be defined by $q_{(u, \alpha)}(s, t) = (u(s), \alpha(s) \circ t)$. The functor $\pi$ sends morphisms $(u, \alpha)$ to squares:
\[
\ti
\{(s, t) \; | \; s \in S, t : \top \to f(s) \} \ar[r, "q_{(u, \alpha)}"] \ar[d, "fst"]& \{(s', t') \; | \; s' \in S', t' : \top \to g(s') \} \ar[d, "fst"]\\
I \ar[r, "u"] & J
\kz
\]
in $\mathbf{Set}$.
\\

The intuitionistic part of our semantic structure will simply be the fibration $p : Fam(\mathbf{Set}) \to \mathbf{Set}$ with $\pi : Fam(\mathbf{Set}) \to \mathbf{Set}^\to$ as outlined above. For the linear part, we choose any symmetric monoidal category $\mathcal{V}$ and form the fibration $q : Fam(\mathcal{V}) \to \mathbf{Set}$. With the functor $I : \mathbf{Set} \to Fam(\mathcal{V})$ defined by mapping a set $S$ to the family constant at the unit of $\mathcal{V}$ and $\otimes : Fam(\mathcal{V}) \times_{\mathbf{Set}} Fam(\mathcal{V}) \to Fam(\mathcal{V})$ induced by the monoidal structure of $\mathcal{V}$, this forms a lax monoidal functor.


We have the following picture:
\[
\ti
Fam(\mathcal{V}) \ar[rd, "q"]  & Fam(\mathbf{Set}) \ar[d, "p"] \ar[r, "\pi"] & \mathbf{Set}^\to \ar[ld, "cod"] \\
& \mathbf{Set}
\kz
\]

In this model, consider the following construction:
\[
\begin{split}
&\vdash A \text{ type}\\
x, y : A &\vdash Arr(x, y) \text{ linear}\\
x, y, z : A &\vdash \_ \circ \_ : Arr(y, z) \otimes Arr(x, y) \multimap Arr(x, z)\\
&and so on
\end{split}
\]
Then $Arr(x, y)$ will be interpreted as a functor $[[Arr(x, y)]] : Free([[A]]) \to \mathcal{V}$. But this is not quite what we want? We want $Arr(x,y)$ to be an object of $\mathcal{V}$.
\subsection{Diagrams}
Expanding upon the Monoidal Families example given by Vakar \cite{vakar14}, we consider diagrams $J : \mathcal{D} \to \mathcal{V}$ of any shape in a category $\mathcal{V}$. In other words, we construct a category $\diag(\mathcal{V})$ whose objects are functors with $\mathcal{V}$ as codomain and whose morphisms for diagrams $J : \mathcal{D} \to \mathcal{V}$ and $J' :\mathcal{C} \to \mathcal{V}$ are given by functors $F : \mathcal{C} \to \mathcal{D}$ such that $J \circ F = J'$.

From this we get the obvious domain fibration $dom : \diag(\mathcal{V}) \to \textbf{Cat}$.
\begin{thm}
If $\mathcal{V}$ is a monoidal category, $dom$ is a lax monoidal fibration.
\begin{proof}
\end{proof}
\end{thm}
This suggests that we can get a model for linear dependent type theory by letting $\mathcal{L} = \diag(\mathcal{V})$ in the semantic structures given above. But what about the choice of $\mathcal{T}$? One can trivially choose $\mathcal{T} = \mathbf{Cat}$ to get the following picture:
\[
\ti
\diag(\mathcal{V}) \ar[rd, "dom"]  & \mathbf{Cat} \ar[d, "1_{\mathbf{Cat}}"] \ar[r, "Id"] & \mathbf{Cat}^\to \ar[ld, "cod"] \\
& \mathbf{Cat}
\kz
\]
where $Id : \mathbf{Cat} \to \mathbf{Cat}^\to$ is the map sending a category to its identity functor. Another choice would be $\mathcal{T} = \mathbf{Fib}$, the category of grothendieck fibrations, yielding:
\[
\ti
\diag(\mathcal{V}) \ar[rd, "dom"]  & \mathbf{Fib} \ar[d, "Base"] \ar[r, "\mathcal{P}"] & \mathbf{Cat}^\to \ar[ld, "cod"] \\
& \mathbf{Cat}
\kz
\]
where $Base$ is the morphism sending a fibration $p : E \to B \mapsto B$ and $\mathcal{P}$ sends a fibration $p : E \to B$ to its underlying functor $p : E \to B$. For any functor $F : A \to B$ in $\mathbf{Cat}$, there is a pullback:
\[
\ti
A \times_{B} E \ar[r] \ar[d, "F^*(p)"] & E \ar[d, "p"] \\
A \ar[r, "F"] & B
\kz
\]
where $F^*(p)$ is a fibration if $p$ is, by Lemma(Prop 2.6 in comprehension categories). This shows that $\mathcal{P}$ sends cartesian arrows to cartesian arrows. READ EXAMPLE 4.6 in Comprehension cats

One could also restrict $\diag(\mathcal{V})$ to diagrams in $\mathcal{V}$ over groupoids. In that case, $\mathcal{T}$ consists of fibrations of groupoids and the intuitionistic fragment of the type theory is essentially the groupoid model as presented by Hofmann and Streicher in \cite{hofmann1998}. This suggests that this model can serve as a setting for linear dependent homotopy type theory.

\newpage
We expand this to a functor $\text{Diag}(\mathcal{V}) : \mathbf{Cat}^{op} \to \mathbf{SMCat(-,V)}$ where $\mathcal{C} \xrightarrow{F} \mathcal{D}$ in \textbf{Cat} induces a monoidal functor $[\mathcal{D}, \mathcal{V}] \xrightarrow{(-) \circ F} [\mathcal{C}, \mathcal{V}]$. To highlight the connection to the type theory, we let $A\{F\}$ in  $\diag(\mathcal{V})(\mathcal{C})$ denote the image of any object $A \in \diag(\mathcal{V})(\mathcal{D})$ under this functor. Notice that for any category $\mathcal{C}$, the functor category $[\mathcal{C}, \mathcal{V}]$ naturally carries a monoidal structure by $(F \otimes G)(c) = F(c) \otimes G(c)$, with the constant functor at the unit object $I \in \mathcal{\Ob(V)}$ as unit. This functor will also be denoted $I$.
\begin{thm}[Comprehension in $\diag(\mathcal{V})$]
$Diag(\mathcal{V})$ admits a comprehension. In other words, for any diagram $J : \mathcal{D} \to \mathcal{V}$ there is a representating object $\mathcal{D}.J \xrightarrow{\mathbf{p}_{\mathcal{D}.J}} \mathcal{D}$ in $\mathbf{Cat}_{/\mathcal{D}}$ such that for any morphism $F : \mathcal{C} \to \mathcal{D}$, we have $\diag(\mathcal{V})(\mathcal{C})(I, J\{F\}) \cong \mathbf{Cat}_{/\mathcal{D}}(F, \mathbf{p}_{(\mathcal{D}, J)})$.
\begin{proof}
Let $\mathcal{D}.J$ be given by the slice category $(I \downarrow J)$, whose objects are pairs $(d, j)$, with $d \in \mathcal{D}$, $j \in \Hom(I,J(d))$ and whose morphisms are $f : d \to d'$ such the following diagram commutes:
\[
\ti
I \ar[r, "j"] \ar[rd, "j'"] & J(d) \ar[d,"J(f)"]\\
& J(d')
\kz
\]
Compare this with natural transformations $\eta \in \diag(\mathcal{V})(\mathcal{C})(I, J\{F\}))$. For all objects $c, c' \in \mathcal{C}$ with morphism $c \xrightarrow{f} c'$ we require components $n_{\bullet} : I \to J \circ F(\bullet)$ such that the following diagram commutes:
\[
\ti
I \ar[r, "\eta_c"] \ar[rd, "\eta_{c'}"] & J(F(c)) \ar[d,"J(F(f))"]\\
& J(F(c'))
\kz
\]
Let $p_{\mathcal{D}.J} : (I \downarrow J) \to \mathcal{D}$ be the obvious forgetful functor, $(d, j) \mapsto d$. Now any functor $G : \mathcal{C} \to (I \downarrow J)$ such that $p_{\mathcal{D}.J}G = F$, must be of the form:
\[
\begin{split}
G(c) &= (F(c), g_c)\\
G(f) &= F(f)
\end{split}
\]
where $g_c : I \to F(c)$ is some morphism satisfying the commutativity conditions above. This uniqely defines a natural transformation $g \in \diag(\mathcal{V})(\mathcal{C})(I, J\{F\})$. Furthermore, for any natural transformation $\eta \in \diag(\mathcal{V})(\mathcal{C})(I, J\{F\}$, there is a corresponding functor $\hat \eta : \mathcal{C} \to (I \downarrow J)$ mapping $c \mapsto (F(c), \eta_c)$.
\end{proof}
\begin{lem}
For any groupoid $G$ and functor $A \in [G, \mathcal{V}]$ the comma category $(I \downarrow A)$ is a groupoid, and its associated projection $p_{G.A} : (I \downarrow A) \to G$ is a fibration.
\begin{proof}
A morphism between objects $(g, i)$, $(g', i')$ in $(I \downarrow A)$ are given by a morphism $f : g \to g'$ in $G$ such that $A(f) \circ i = i'$. Since $G$ is a groupoid there exists an inverse $f^{-1}$. That this map yields an inverse of $(I \downarrow A)$ is immediate from the fact that the following diagram in $\mathcal{V}$ commutes:
\[
\ti
I \ar[rd,"i'",swap] \ar[r, "i"] & A(g) \ar[d, "A(f)", bend right,swap] \\
& A(g') \ar[u, "A(f^{-1}) = A(f)^{1}", bend right,swap]
\kz
\]
To see that $p$ is a fibration, let $(g, i)$ be an object of $(I \downarrow A)$ and $f : g \to g'$ an arrow in $G$. Then the object $(g', A(f)\circ i)$ together with $f$ seen as a morphism of $(I \downarrow A)$ form an immediate commutative diagram in $\mathcal{V}$.
\end{proof}
\end{lem}
\begin{cor}
The restriction of $\diag(\mathcal{V})$ to the category of groupoids is also a model for IDLTT.
\end{cor}

\end{thm}
The functor $p_{\mathcal{D}.J}$ will sometimes be referred to as the \textbf{projection} associated to $J$.
Recall that a model $\mathcal{C}, \mathcal{L}$ of ILDTT \textbf{supports $\Sigma$-types} if $\mathcal{L}(p_{\Delta.A})$ has a left adjoint for all $\Delta \in \mathcal{C}$, and $A \in \mathcal{L}(\Delta)$. It \textbf{supports $\Pi$-types} if all $\mathcal{L}(p_{\Delta.A})$ have right adjoints.
\begin{rem}
In the $Diag(V)$ model, $\Sigma$- and $\Pi$-types are left and right adjoints to the functor $p_* : [\Delta, \mathcal{V}] \to [\Delta.A, \mathcal{V}]$ induced by the projection $p_{\Delta.A} : \Delta.A \to \Delta$. These are precisely the left and right Kan extensions along $p_{\Delta.A}$.
\end{rem}
\begin{defn}[coCartesian morphism]
Given a functor $p : E \to B$, an arrow $f : e_1 \to e_2$ is coCartesian with respect to $p$ if, for any pair of morphisms $h : e_1 \to e_3$ and $g : p(e_2) \to p(e_3)$ as in the following commutative diagram:
\[
\ti
p(e_1) \ar[r, "p(f)"] \ar[rd, "p(h)"] & p(e_2) \ar[d, "g"] \\
& p(e_3)
\kz
\]
there exists a unique morphism $\hat g : e_2 \to e_3$ such that $p(\hat g) = g$ and $\hat g f = h$.
\end{defn}
\begin{lem}{unique-target}
For a functor $p : E \to B$ and $f_1 : e_1 \to e_2$, $f_2 : e_1 \to e_3$ two coCartesian morphisms such that $p(f_1) = p(f_2) = \alpha : b_1 \to b_2$, there is a unique isomorphisms $\phi : e_2 \to e_3$ such that the following diagram commutes:
\[
\ti
e_1 \ar[d, "f_1"] \ar[rd, "f_2"] & \\
e_2 \ar[r,"\phi"] &  e_3
\kz
\]
\begin{proof}
Since $f_1$ coCartesian, the composition $Id_{b_2} \circ p(f_1) = p(f_2)$ has a unique lift $\phi : e_2 \to e_3$ in $E$ such that $\phi f_1 = f_2$. Similarly, there is a unqie morphism $\psi : e_3 \to e_2$ such that $\psi f_2 = f_1$. Composing, we get an arrow $\psi \phi f_1 = f_1$. But since we also have $Id_{b_2} \circ p(f_1) = p(f_1)$, the only arrow $\pi : e_2 \to e_2$ such that $\pi f_1 = f_1$ must be the identity on $e_2$, so $\phi$ is a isomorphism.
\end{proof}
\end{lem}
\begin{defn}[Opfibration]
A functor $p : E \to B$ is an \textbf{opfibration} if for every $e_1 \in E$ and arrow $\alpha : p(e_1) \to b$ in the base $B$, there exists a coCartesian arrow $f : e_1 \to e_2$ in $E$ such that $p(f) = \alpha$.
\end{defn}
Given such a fibration, for any object $b \in B$, the objects and morphisms of $E$ that are mapped to $b$ and its identity morphism via $p$ form a category, $E_b$, called the $\textbf{fiber}$ over $b$. From any map $f : b \to b'$ in the base we can construct a functor $\hat f : E_b \to E_{b'}$ by sending $e \in E_b$ to the target of a corresponding coCartesian lift $g : e \to e'$. By the previous lemma, such a functor $\hat f$ will be unique up to unique isomorphism.
\begin{thm}
Given a category $\mathcal{C}$ and functor $A : \mathcal{C} \to \mathcal{V}$, the projection $p : (v \downarrow A) \to \mathcal{C}$ is an opfibration for any $v \in \mathcal{V}$.
\begin{proof}
Let $(c, j: v \to A(c))$ be an object of $(v \downarrow A)$ and $f: c \to c'$ a morphism in $\mathcal{C}$. Then $f$ induces a morphism of between $(c, j)$ and $(c', fj)$ in $(v \downarrow A)$. For any pair of morphisms $h : (c, j) \to (c'', j'')$ and $g : c' \to c''$ such that the following diagram commutes:
\[
\ti
c \ar[r, "f"] \ar[rd, "h"] & c' \ar[d, "g"]\\
& c''
\kz
\]
we have the following commutative diagram in $(v \downarrow A)$:
\[
\ti
& c \ar[d, "f"] \ar[dd, "h", bend left]\\
v \ar[ru, "j"] \ar[rd, "j''"] \ar[r, "f j'"]& c' \ar[d, dotted, "g"]\\
& c''
\kz
\]
Furthermore, since morphisms in $(v \downarrow A)$ arise from morphisms in $\mathcal{C}$, $g$ is the unique candidate.
\end{proof}
\end{thm}
\begin{thm}
Let $p : \mathcal{C} \to \mathcal{D}$ be an opfibration and $\mathcal{V}$  cocomplete. Then there exists a functor $Lan_p : [\mathcal{C}, \mathcal{V}] \to [\mathcal{D}, \mathcal{V}]$ which is a left kan extension of $p$.
\begin{proof}
For any functor $Y \in [\mathcal{C}, \mathcal{V}]$, let $Lan_p(Y)$ be given by:
\[
\begin{split}
Lan_p(Y)(d) &:= \lim_{\to}(\mathcal{C}_d \hookrightarrow \mathcal{C} \xrightarrow{Y} \mathcal{V})\\
\end{split}
\]
The action of $Lan_p(Y)$ on morphisms $f : d \to d'$, arises from the universal mapping property of the colimits. Specifically, one can distinguish between two cases. If $d$ is not in the image of $p$, then $Lan_p(Y)(d)$ is the initial object in $\mathcal{V}$ and $Lan_p(Y)(f)$ the unique morphism from it. Otherwise, let $p(c) = d$ for some $c \in \mathcal{C}$ and $\hat f : c \to c'$ be a cocartesian arrow such that $p(\hat f) = f$. We want to show that $Lan_p(Y)(d')$ forms a cocone of the diagram to which $Lan_p(Y)(d)$ is a limit. To that end, let $g : c \to c''$ be an arbitrary morphism in $\mathcal{C}_{d}$. We have the following image:
\[
\ti
Y(c) \ar[r, "\lambda_{Y(c)}"] \ar[dd, "Y(\hat f)",swap, bend right] \ar[d, "Y(g)"]  & Lan_p(Y)(d) \\
Y(c'')  \ar[ru, "\lambda_{Y(c')}"]  \\
Y(c') \ar[r, "\lambda_{Y(c'')}", swap] & Lan_p(Y)(d') \
\kz
\]
But there should also exist a cocartesian arrow $\hat g$ corresponding to $1_d : p(c) \to p(c'')$, so from the following diagram in $\mathcal{D}$:
\[
\ti
p(c) = d \ar[d, "p(\hat f) = f",swap] \ar[r, "p(\hat g) = 1_d"] & p(c'') = d\ar[ld, "f"] \\
p(c') = d'
\kz
\]
there must exist a unique arrow $f' : c'' \to c'$ such that $p(f') = f$ and $f' \hat g = \hat f$. This implies that $Lan_p(Y)(d')$ forms a cocone over the diagram in question, so we can define the $Lan_p(Y)(f)$ to be the unique map $Lan_p(Y)(d) \to Lan_p(Y)(d')$.
The action of $Lan_p$ on natural transformations $\epsilon : Y \to Z$ is similarly induced by the UMP of the colimits at each component.
\\
We show  $Lan_p \dashv p_*$ by exhibiting the unit $\eta :  1_{[\mathcal{D}, \mathcal{V}]} \to Lan_pp_*$ with the following universal property. For objects $Y \in [\mathcal{C}, \mathcal{V}]$, $X \in [\mathcal{D}, \mathcal{V}]$ and morphism $f : Y \to X \circ p$ there is a unique morphism $g : Lan_p(Y) \to X$  such that the following diagram commutes:
\[
\ti
Y \ar[d, "\eta_Y"] \ar[rd, "f"]\\
Lan_p(Y) \circ p \ar[r, "p*(g)"] &  X \circ p
\kz
\]
Again, $g$ arises from the universal property of the colimit: For a given $d \in \mathcal{D}$, all $c \in \mathcal{C}$ such that $p(c) = d$ are mapped to the same object by $X \circ p$, so $X \circ p(c)$ forms a cocone over the diagram $\mathcal{C}_{p(c)} \hookrightarrow \mathcal{C} \xrightarrow{Y} \mathcal{V}$. But $Lan_p(Y)(p(c))$ is the colimit of that diagram, so the $g$ arises from the unique map from the colimit to the cocone.
\end{proof}
\end{thm}
Dually, we can use the same argument to conclude that when $\mathcal{V}$ has all limits, the right Kan extension along $p$ exists.
\begin{cor}
If $\mathcal{V}$ has all colimits, then $\diag(\mathcal{V})$ supports $\Sigma$-types. If $\mathcal{V}$ has all limits, then $\diag(\mathcal{V})$ supports $\Pi$-types.
\end{cor}
\subsection{Spectra}
%\subsection{Monoidal groupoids}
%Given any bicategory $\mathcal{B}$, there is a monoidal groupoid $End(b)$ of endomorphisms of an object $b \in \mathcal{B}$ with 1-cells $f : b \to b$ as objects and invertible 2-cells $\eta : f \implies f'$ as morphisms. The monoidal structure of $End(b)$ is given by composition: $f \otimes f' = f'\circ f$. In particular, if $\mathcal{B} = \mathbf{Cat}$, then we have $End(\mathcal{C})$ a monoidal groupoid consisting of functors and natural isomorphisms.\\

\newpage


\section{Ideas}
H-spaces generalized to the tensor product?
\subsection{Dependent ordered type theory}
If one were to drop exchange as a structural rule, this should correspond to dropping the requirement of the semantic category being symmetric. A particular construction which might be interesting to explore in this setting would be simplicial sets equipped with the join. Would this work?
\bibliographystyle{plain}
\bibliography{refs.bib}
\end{document}

